---
title: "- Exploratory Data Analysis -"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

<style>
body{text-align: justify}
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

*File creation: October, 21st 2021*  
*Update: July, 27th 2022*  

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(Seurat)
library(MAST)
library(doMC)
library(cowplot)
library(data.table)
# library(slingshot)
# library(monocle3)
```
```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
load(here("output", "002_output", "002_EDA.RData"))
# save.image(here("output", "002_output", "002_EDA.RData"))
```

# Exploratory Data Analysis - CD4+ T
____________________________________

Two ADTs (CD4 and CD8) are used to determine CD4+ and CD8+ T cells using the `SCAMP` R package. Once CD4+ T cells are extracted, we re-run the WNN workflow - see code below.  
This workflow consists of three steps: 1) independent preprocessing and dimensional reduction of each modality individually; 2) learning cell-specific modality *weights*, and constructing a WNN graph that integrates the modalities; and 3) downstream analysis (i.e. visualization, clustering, etc.) of the WNN graph. The `FindMultiModalNeighbors` R function is run with `dims.list = list(1:50, 1:25)` and `modality.weight.name = c("RNA.weight", "ADT.weight")`. Before re-running the WNN workflow, PCA (ADT & GEX) are also re-run in order to avoid having potential PCs driven by other factors (AgSpecNot or cell types).  

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- CD4+ T - AgSpec - GEX
#- Subset
integrated.seurat <- readRDS(file = here("data", "001_integrated-seurat_GEX.rds"))
integrated.seurat <- subset(integrated.seurat, sort_population == "AgSpec")
integrated.seurat <- subset(integrated.seurat, T_phenotype == "CD4+")
DefaultAssay(integrated.seurat)

#- Data Integration done (w/ AgSpec & AgSpecNot; w/ CD4+ T, CD8+ T & DURTs)

#- Normalization
integrated.seurat <- NormalizeData(object = integrated.seurat, assay = "RNA", normalization.method = "LogNormalize", verbose = TRUE)
integrated.seurat <- NormalizeData(object = integrated.seurat, assay = "ADT", normalization.method = "CLR", verbose = TRUE)

#- PCA
VariableFeatures(integrated.seurat) <- VariableFeatures(integrated.seurat)[which(!str_detect(string = VariableFeatures(integrated.seurat), pattern = "^TRAJ"))]
VariableFeatures(integrated.seurat) <- VariableFeatures(integrated.seurat)[which(!str_detect(string = VariableFeatures(integrated.seurat), pattern = "^TRBJ"))]
VariableFeatures(integrated.seurat) <- VariableFeatures(integrated.seurat)[which(!str_detect(string = VariableFeatures(integrated.seurat), pattern = "^TRGJ"))]
VariableFeatures(integrated.seurat) <- VariableFeatures(integrated.seurat)[which(!str_detect(string = VariableFeatures(integrated.seurat), pattern = "^TRDJ"))]
VariableFeatures(integrated.seurat) <- VariableFeatures(integrated.seurat)[which(!str_detect(string = VariableFeatures(integrated.seurat), pattern = "^TRAD"))]
VariableFeatures(integrated.seurat) <- VariableFeatures(integrated.seurat)[which(!str_detect(string = VariableFeatures(integrated.seurat), pattern = "^TRBD"))]
VariableFeatures(integrated.seurat) <- VariableFeatures(integrated.seurat)[which(!str_detect(string = VariableFeatures(integrated.seurat), pattern = "^TRGD"))]
VariableFeatures(integrated.seurat) <- VariableFeatures(integrated.seurat)[which(!str_detect(string = VariableFeatures(integrated.seurat), pattern = "^TRDD"))]
VariableFeatures(integrated.seurat) <- VariableFeatures(integrated.seurat)[which(!str_detect(string = VariableFeatures(integrated.seurat), pattern = "^TRAV"))]
VariableFeatures(integrated.seurat) <- VariableFeatures(integrated.seurat)[which(!str_detect(string = VariableFeatures(integrated.seurat), pattern = "^TRBV"))]
VariableFeatures(integrated.seurat) <- VariableFeatures(integrated.seurat)[which(!str_detect(string = VariableFeatures(integrated.seurat), pattern = "^TRGV"))]
VariableFeatures(integrated.seurat) <- VariableFeatures(integrated.seurat)[which(!str_detect(string = VariableFeatures(integrated.seurat), pattern = "^TRDV"))]
integrated.seurat <- RunPCA(object = integrated.seurat, npcs = 100, verbose = TRUE)
ElbowPlot(integrated.seurat, ndims = 100)
chosen.nPCs <- 50

#- UMAP
integrated.seurat <- RunUMAP(object = integrated.seurat, dims = 1:chosen.nPCs, verbose = TRUE, seed.use = 1234, min.dist = .01)

#- Output
saveRDS(object = integrated.seurat, file = here("data", "002_integrated-seurat_GEX.rds"))
```
```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- CD4+ T - AgSpec - ADT
#- Subset
integrated.seurat <- readRDS(file = here("data", "001_integrated-seurat_ADT.rds"))
integrated.seurat <- subset(integrated.seurat, sort_population == "AgSpec")
integrated.seurat <- subset(integrated.seurat, T_phenotype == "CD4+")
DefaultAssay(integrated.seurat)

#- Data Integration done (w/ AgSpec & AgSpecNot; w/ CD4+ T, CD8+ T & DURTs)

#- Normalization
integrated.seurat <- NormalizeData(object = integrated.seurat, assay = "RNA", normalization.method = "LogNormalize", verbose = TRUE)
integrated.seurat <- NormalizeData(object = integrated.seurat, assay = "ADT", normalization.method = "CLR", verbose = TRUE)

#- Scaling & PCA
integrated.seurat <- ScaleData(object = integrated.seurat, vars.to.regress = "nCount_ADT",verbose = TRUE)
integrated.seurat <- RunPCA(object = integrated.seurat, npcs = 100, verbose = TRUE)
ElbowPlot(integrated.seurat, ndims = 100)
chosen.nPCs <- 25

#- UMAP
integrated.seurat <- RunUMAP(object = integrated.seurat, dims = 1:chosen.nPCs, verbose = TRUE, seed.use = 1234, min.dist = .01)

#- Output
saveRDS(object = integrated.seurat, file = here("data", "002_integrated-seurat_ADT.rds"))
```
```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- CD4+ T - AgSpec - WNN
#- Data
integrated.seurat_GEX <- readRDS(file = here("data", "002_integrated-seurat_GEX.rds"))
#ElbowPlot(integrated.seurat_GEX, ndims = 100) # 50 PCs
integrated.seurat_ADT <- readRDS(file = here("data", "002_integrated-seurat_ADT.rds"))
#ElbowPlot(integrated.seurat_ADT, ndims = 100) # 50 PCs

#- Add PCA results to one Seurat object
integrated.seurat_GEX[["pca_ADT"]] <- CreateDimReducObject(embeddings = Embeddings(object = integrated.seurat_ADT, reduction = "pca"), assay = "integrated")

#- Run WNN
integrated.seurat_WNN <- FindMultiModalNeighbors(object = integrated.seurat_GEX,
                                                 reduction.list = list("pca", "pca_ADT"),
                                                 dims.list = list(1:50, 1:25),
                                                 modality.weight.name = c("RNA.weight", "ADT.weight"),
                                                 prune.SNN = 1/20)

#- UMAP
integrated.seurat_WNN <- RunUMAP(object = integrated.seurat_WNN, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_", seed.use = 1234, min.dist = .1, n.neighbors = 15, verbose = TRUE)

#- Clustering
set.seed(1234)
resolutions <- c(seq(from = .1, to = 1, by = .1), seq(from = 1.2, to = 2, by = .2))
for(i in resolutions) {
  integrated.seurat_WNN <- FindClusters(object = integrated.seurat_WNN, graph.name = "wsnn", algorithm = 3, resolution = i, n.start = 10, verbose = TRUE)
}

#- Output
saveRDS(object = integrated.seurat_WNN, file = here("data", "002_integrated-seurat_WNN.rds"))
```
```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- Processing - GEX + ADT = WNN
#- Open .rds
integrated.seurat_GEX <- readRDS(file = here("data", "002_integrated-seurat_GEX.rds"))
integrated.seurat_ADT <- readRDS(file = here("data", "002_integrated-seurat_ADT.rds"))
integrated.seurat_WNN <- readRDS(file = here("data", "002_integrated-seurat_WNN.rds"))
identical(colnames(integrated.seurat_WNN), colnames(integrated.seurat_GEX))
identical(colnames(integrated.seurat_WNN), colnames(integrated.seurat_ADT))

#- DietSeurat
DefaultAssay(integrated.seurat_WNN) <- "RNA"
reducedSize_integrated.seurat <- DietSeurat(object = integrated.seurat_WNN, 
                                            counts = TRUE, 
                                            data = TRUE, 
                                            assays = c("RNA", "ADT"), 
                                            dimreducs = NULL)

#- Add UMAPs
reducedSize_integrated.seurat[["umap_GEX"]] <- CreateDimReducObject(embeddings = Embeddings(object = integrated.seurat_GEX, reduction = "umap"), assay = "RNA")
reducedSize_integrated.seurat[["umap_ADT"]] <- CreateDimReducObject(embeddings = Embeddings(object = integrated.seurat_ADT, reduction = "umap"), assay = "ADT")
reducedSize_integrated.seurat[["umap_WNN"]] <- CreateDimReducObject(embeddings = Embeddings(object = integrated.seurat_WNN, reduction = "wnn.umap"), assay = "RNA")

#- Add meta.data
colnames(reducedSize_integrated.seurat@meta.data)
reducedSize_integrated.seurat@meta.data <- reducedSize_integrated.seurat@meta.data[, c(1:3, 31, 4:17, 33:34, 20, 27, 35:36, 37, 39:52)]

#- Output
saveRDS(object = reducedSize_integrated.seurat, file = here("data", "002_integrated-seurat.rds"))
```

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
###--- Open .rds
integrated.seurat <- readRDS(file = here("data", "002_integrated-seurat.rds"))
# saveRDS(object = integrated.seurat, file = here("data", "002_integrated-seurat.rds"))
```

## UMAP - Technical factors

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=12, fig.height=10}
###--- UMAP
data.table(Embeddings(object = integrated.seurat, "umap_WNN"),
           integrated.seurat@meta.data) %>%
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
    geom_point(aes(color = nCount_RNA), size = .01) +
    scale_color_viridis_c(option = "C") +
    labs(x = "UMAP-1", y = "UMAP-2", color = "# UMIs\nGEX") +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 14),
          panel.grid = element_blank(),
          legend.text = element_text(size = 12),
          legend.title = element_text(size = 14)) -> plotA
data.table(Embeddings(object = integrated.seurat, "umap_WNN"),
           integrated.seurat@meta.data) %>%
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
    geom_point(aes(color = nFeature_RNA), size = .01) +
    scale_color_viridis_c(option = "C") +
    labs(x = "UMAP-1", y = "UMAP-2", color = "# genes") +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 14),
          panel.grid = element_blank(),
          legend.text = element_text(size = 12),
          legend.title = element_text(size = 14)) -> plotB
data.table(Embeddings(object = integrated.seurat, "umap_WNN"),
           integrated.seurat@meta.data) %>%
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
    geom_point(aes(color = percent.mt), size = .01) +
    scale_color_viridis_c(option = "C") +
    labs(x = "UMAP-1", y = "UMAP-2", color = "% mito") +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 14),
          panel.grid = element_blank(),
          legend.text = element_text(size = 12),
          legend.title = element_text(size = 14)) -> plotC
data.table(Embeddings(object = integrated.seurat, "umap_WNN"),
           integrated.seurat@meta.data) %>%
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
    geom_point(aes(color = nCount_ADT), size = .01) +
    scale_color_viridis_c(option = "C") +
    labs(x = "UMAP-1", y = "UMAP-2", color = "# UMIs\nADT") +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 14),
          panel.grid = element_blank(),
          legend.text = element_text(size = 12),
          legend.title = element_text(size = 14)) -> plotD
cowplot::plot_grid(plotA, plotB, plotC, plotD, labels = "AUTO", align = c("hv"), ncol = 2)  
```

## UMAP - Batch  

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=6, fig.height=7}
###--- UMAP
data.table(Embeddings(object = integrated.seurat, "umap_WNN"),
           integrated.seurat@meta.data) %>%
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
    geom_point(aes(color = batch), size = .01) +
    labs(x = "UMAP-1", y = "UMAP-2", color = NULL) +
    guides(colour = guide_legend(override.aes = list(size = 4))) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 14),
          panel.grid = element_blank(),
          legend.text = element_text(size = 8),
          legend.position = "bottom")
```
```{r echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
###--- UMAP
data.ggplot <- data.table(Embeddings(object = integrated.seurat, "umap_WNN"), integrated.seurat@meta.data)
for(j in data.ggplot$batch %>% unique()){
  data.ggplot %>%
    ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
      geom_point(color = "grey80", size = .01) +
      ggpointdensity::geom_pointdensity(aes(x = wnnUMAP_1, y = wnnUMAP_2), data = subset(data.ggplot, batch == j), size = .01) +
      viridis::scale_color_viridis() +
      labs(x = "UMAP-1", y = "UMAP-2", color = NULL, title = paste("Summary: Batch", j)) +
      theme_bw() +
      theme(axis.text = element_blank(),
            axis.ticks = element_blank(),
            axis.title = element_text(size = 14),
            panel.grid = element_blank(),
            legend.text = element_text(size = 6),
            legend.position = "none") -> plot
  png(here::here("output", "002_output", paste0("WNN_UMAP-by-batch_", j, ".png")), width = 500, height = 550, res = 120)
  print(plot)
  dev.off()
}
```

## UMAP - PTID

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=6, fig.height=7}
###--- UMAP
data.table(Embeddings(object = integrated.seurat, "umap_WNN"),
           integrated.seurat@meta.data) %>%
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
    geom_point(aes(color = ptid %>% as.character), size = .01) +
    labs(x = "UMAP-1", y = "UMAP-2", color = NULL) +
    guides(colour = guide_legend(override.aes = list(size = 4))) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 14),
          panel.grid = element_blank(),
          legend.text = element_text(size = 6),
          legend.position = "bottom")
```
```{r echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
###--- UMAP
data.ggplot <- data.table(Embeddings(object = integrated.seurat, "umap_WNN"), integrated.seurat@meta.data) 
for(j in data.ggplot$ptid %>% as.character() %>% unique()){
  data.ggplot %>%
    ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
      geom_point(color = "grey80", size = .01) +
      ggpointdensity::geom_pointdensity(aes(x = wnnUMAP_1, y = wnnUMAP_2), data = subset(data.ggplot, ptid == j), size = .01) +
      viridis::scale_color_viridis() +
      labs(x = "UMAP-1", y = "UMAP-2", color = NULL, title = paste("Summary: PTID", j)) +
      theme_bw() +
      theme(axis.text = element_blank(),
            axis.ticks = element_blank(),
            axis.title = element_text(size = 14),
            panel.grid = element_blank(),
            legend.text = element_text(size = 6),
            legend.position = "none") -> plot
  png(here::here("output", "002_output", paste0("WNN_UMAP-by-ptid_", j, ".png")), width = 500, height = 550, res = 120)
  print(plot)
  dev.off()
}
```

## UMAP - Arm

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
###--- UMAP
data.table(Embeddings(object = integrated.seurat, "umap_WNN"),
           integrated.seurat@meta.data) %>%
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
    geom_point(aes(color = arm), size = .01) +
    scale_color_brewer(palette = "Dark2") +
    labs(x = "UMAP-1", y = "UMAP-2", color = NULL) +
    facet_wrap(~ arm, ncol = 2) +
    guides(colour = guide_legend(override.aes = list(size = 4))) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 14),
          panel.grid = element_blank(),
          legend.text = element_text(size = 12),
          legend.position = "bottom")
```
```{r echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
###--- UMAP
data.ggplot <- data.table(Embeddings(object = integrated.seurat, "umap_WNN"), integrated.seurat@meta.data) 
for(j in data.ggplot$arm %>% unique()){
  data.ggplot %>%
    ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
      geom_point(color = "grey80", size = .01) +
      ggpointdensity::geom_pointdensity(aes(x = wnnUMAP_1, y = wnnUMAP_2), data = subset(data.ggplot, arm == j), size = .01) +
      viridis::scale_color_viridis() +
      labs(x = "UMAP-1", y = "UMAP-2", color = NULL, title = paste("Summary: Arm", j)) +
      theme_bw() +
      theme(axis.text = element_blank(),
            axis.ticks = element_blank(),
            axis.title = element_text(size = 14),
            panel.grid = element_blank(),
            legend.text = element_text(size = 6),
            legend.position = "none") -> plot
  png(here::here("output", "002_output", paste0("WNN_UMAP-by-arm_", j, ".png")), width = 500, height = 550, res = 120)
  print(plot)
  dev.off()
}
```

## UMAP - Stimulation

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
###--- UMAP
data.table(Embeddings(object = integrated.seurat, "umap_WNN"),
           integrated.seurat@meta.data) %>%
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
    geom_point(aes(color = stim), size = .01) +
    scale_color_brewer(palette = "Set3", direction = -1) +
    labs(x = "UMAP-1", y = "UMAP-2", color = NULL) +
    facet_wrap(~ stim, ncol = 2) +
    guides(colour = guide_legend(override.aes = list(size = 4))) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 14),
          panel.grid = element_blank(),
          legend.text = element_text(size = 12),
          legend.position = "bottom")
```
```{r echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
###--- UMAP
data.ggplot <- data.table(Embeddings(object = integrated.seurat, "umap_WNN"), integrated.seurat@meta.data) 
for(j in data.ggplot$stim %>% unique()){
  data.ggplot %>%
    ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
      geom_point(color = "grey80", size = .01) +
      ggpointdensity::geom_pointdensity(aes(x = wnnUMAP_1, y = wnnUMAP_2), data = subset(data.ggplot, stim == j), size = .01) +
      viridis::scale_color_viridis() +
      labs(x = "UMAP-1", y = "UMAP-2", color = NULL, title = paste("Summary: Stimulation", j)) +
      theme_bw() +
      theme(axis.text = element_blank(),
            axis.ticks = element_blank(),
            axis.title = element_text(size = 14),
            panel.grid = element_blank(),
            legend.text = element_text(size = 6),
            legend.position = "none") -> plot
  png(here::here("output", "002_output", paste0("WNN_UMAP-by-stimulation_", j, ".png")), width = 500, height = 550, res = 120)
  print(plot)
  dev.off()
}
```

## UMAP - Visit #

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
###--- UMAP
data.table(Embeddings(object = integrated.seurat, "umap_WNN"),
           integrated.seurat@meta.data) %>%
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
    geom_point(aes(color = visit), size = .01) +
    scale_color_brewer(palette = "Paired") +
    labs(x = "UMAP-1", y = "UMAP-2", color = NULL) +
    facet_wrap(~ visit, ncol = 2) +
    guides(colour = guide_legend(override.aes = list(size = 4))) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 14),
          panel.grid = element_blank(),
          legend.text = element_text(size = 12),
          legend.position = "bottom")
```
```{r echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
###--- UMAP
data.ggplot <- data.table(Embeddings(object = integrated.seurat, "umap_WNN"), integrated.seurat@meta.data) 
for(j in data.ggplot$visit %>% unique()){
  data.ggplot %>%
    ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
      geom_point(color = "grey80", size = .01) +
      ggpointdensity::geom_pointdensity(aes(x = wnnUMAP_1, y = wnnUMAP_2), data = subset(data.ggplot, visit == j), size = .01) +
      viridis::scale_color_viridis() +
      labs(x = "UMAP-1", y = "UMAP-2", color = NULL, title = paste("Summary: Visit", j)) +
      theme_bw() +
      theme(axis.text = element_blank(),
            axis.ticks = element_blank(),
            axis.title = element_text(size = 14),
            panel.grid = element_blank(),
            legend.text = element_text(size = 6),
            legend.position = "none") -> plot
  png(here::here("output", "002_output", paste0("WNN_UMAP-by-visitno_", j, ".png")), width = 500, height = 550, res = 120)
  print(plot)
  dev.off()
}
```

## UMAP - Markers

**Markers - GEX**

Not shown.  

**Markers - ADT**

Not shown.  

```{r echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
###--- UMAP
data.ggplot <- data.table(Embeddings(object = integrated.seurat, "umap_WNN"),
                          integrated.seurat@assays$ADT@data %>% as.matrix() %>% t() %>% as.matrix()) %>%
  reshape2::melt(id.vars = c("wnnUMAP_1", "wnnUMAP_2")) 
for(j in data.ggplot$variable %>% unique()){
  data.ggplot %>%
    filter(variable == j) %>%
    mutate(value = ifelse(value > 3, 3, value)) %>%
    ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
      geom_point(aes(color = value), size = .01) +
      scale_color_gradient(low = "lightgrey", high = "darkgreen", breaks = c(0, 1, 2, 3), labels = c("0", "1", "2", ">3")) +
      labs(x = "UMAP-1", y = "UMAP-2", color = "ADT", title = paste("Summary: ADT", j)) +
      theme_bw() +
      theme(axis.text = element_blank(),
            axis.ticks = element_blank(),
            axis.title = element_text(size = 18),
            strip.text = element_text(size = 18),
            panel.grid = element_blank(),
            legend.text = element_text(size = 12),
            legend.position = "bottom") -> plot
  png(here::here("output", "002_output", paste0("WNN_UMAP-by-ADT-marker_", j, ".png")), width = 500, height = 600, res = 120)
  print(plot)
  dev.off()
}
```

## Clustering

With a `resolution = 1`, we have 20 clusters    

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=6, fig.height=6}
###--- UMAP
Idents(integrated.seurat) <- factor(x = integrated.seurat@meta.data$wsnn_res.1, levels = 0:19) 
data.table(Embeddings(object = integrated.seurat, "umap_WNN"), wsnn_res.1 = Idents(integrated.seurat)) %>%
  select(wnnUMAP_1, wnnUMAP_2, wsnn_res.1) %>%
  reshape2::melt(id.vars = c("wnnUMAP_1", "wnnUMAP_2")) %>%
  mutate(resolution = str_replace(string = variable, pattern = "wsnn_res.", replacement = "resolution ")) %>%
  mutate(value = factor(x = value, levels = 0:19)) %>%
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
    geom_point(aes(color = value), size = .01) +
    labs(x = "UMAP-1", y = "UMAP-2", color = NULL) +
    facet_wrap(~ resolution, ncol = 5) +
    guides(colour = guide_legend(override.aes = list(size = 4))) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 14),
          panel.grid = element_blank(),
          legend.text = element_text(size = 12),
          legend.position = "none")
```

When aiming to identify markers of cell state independent of experimental condition, we include batch, donor, arm, stimulation and visit as covariates to the model. To identify gene markers, the `FindAllMarkers` R function is used with `test.use = MAST`, `latent.vars = c("nCount_RNA", "batch", "ptid", "arm", "stimulation", "visit")`, and `max.cells.per.ident = 5000`.  
To identify ADT markers, the `FindAllMarkers` R function is used with `test.use = LR`, `latent.vars = c("batch", "ptid", "arm", "stimulation", "visit")`, and `max.cells.per.ident = 5000`. We merge clusters that do not exhibit clear evidence of separation.  

```{r echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
###--- UMAP
data.table(Embeddings(object = integrated.seurat, "umap_WNN"),
           integrated.seurat@meta.data) %>%
  select(wnnUMAP_1, wnnUMAP_2, wsnn_res.1) %>%
  mutate(wsnn_res.1 = factor(x = wsnn_res.1, levels = 0:19)) %>%
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
    geom_point(aes(color = wsnn_res.1), size = .01) +
    labs(x = "UMAP-1", y = "UMAP-2", color = NULL) +
    facet_wrap(~ wsnn_res.1, ncol = 5) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 14),
          panel.grid = element_blank(),
          legend.text = element_text(size = 12),
          legend.position = "none")
```
```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- Cluster annotation - GEX & ADT
#- Variables
annotations <- read.table(here("misc-docs", "2020_gene-annotation.txt"), header = TRUE, sep = ",")
Idents(integrated.seurat) <- factor(x = integrated.seurat@meta.data$wsnn_res.1, levels = 0:19) 
Idents(integrated.seurat) %>% table()
integrated.seurat@meta.data$batch %>% class()
integrated.seurat@meta.data$ptid %>% class()
integrated.seurat@meta.data$ptid <- integrated.seurat@meta.data$ptid %>% as.character()
integrated.seurat@meta.data$arm %>% class()
integrated.seurat@meta.data$stim %>% class()
integrated.seurat@meta.data$visit %>% class()

#- MAST (GEX)
# Perform differential expression for each of the merged clusters
DefaultAssay(integrated.seurat) <- "RNA"
markers.clusters_GEX <- FindAllMarkers(object = integrated.seurat, 
                                       max.cells.per.ident = 5000,
                                       only.pos = TRUE, 
                                       test.use = "MAST", 
                                       latent.vars = c("nCount_RNA", "batch", "ptid", "arm", "stim", "visit"),
                                       verbose = TRUE)
markers.clusters_GEX <- markers.clusters_GEX %>%
  filter(p_val_adj < 1e-05) %>%
  arrange(cluster, p_val_adj, avg_log2FC) %>%
  left_join(y = unique(annotations[, c("gene_name", "description")]), by = c("gene" = "gene_name")) 
markers.clusters_GEX %>%
  write_csv(file = here("output", "002_output", "GEX_markers-clusters_v1.csv"))

#- Top 20 (GEX)
markers.clusters_GEX %>%
  group_by(cluster) %>% 
  top_n(n = 20, wt = avg_log2FC) %>%
  View("Top genes")

#- LR (ADT)
# Perform differential expression for each of the merged clusters
DefaultAssay(integrated.seurat) <- "ADT"
markers.clusters_ADT <- FindAllMarkers(object = integrated.seurat, 
                                       max.cells.per.ident = 5000,
                                       only.pos = TRUE, 
                                       test.use = "LR", 
                                       latent.vars = c("batch", "ptid", "arm", "stim", "visit"),
                                       verbose = TRUE)
markers.clusters_ADT <- markers.clusters_ADT %>%
  filter(p_val_adj < 1e-05) %>%
  arrange(cluster, p_val_adj, avg_log2FC)
markers.clusters_ADT %>%
  write_csv(file = here("output", "002_output", "ADT_markers-clusters_v1.csv"))

#- Top 20 (ADT)
markers.clusters_ADT %>%
  group_by(cluster) %>% 
  top_n(n = 20, wt = avg_log2FC) %>%
  View("Top ADTs")

#- Ridge plots
pdf(file = here("output", "002_output", "ADT_ridgePlots_v1.pdf"), width = 5, height = 8, bg = "white", pointsize = 12)
for(i in setdiff(rownames(integrated.seurat@assays$ADT@data), "CD4.2")){
  data.table(integrated.seurat@meta.data, ADT = integrated.seurat@assays$ADT@data[i, ]) %>%
    mutate(integrated.clusters = factor(x = wsnn_res.1, levels = 0:19) ) %>%
    ggplot(aes(y = integrated.clusters, x = ADT, fill = integrated.clusters)) +
      ggridges::geom_density_ridges(bandwidth = .2) +
      geom_vline(xintercept = 0, color = "grey20", alpha = .7, lty = 2) +
      labs(x = "Normalized ADT expression", title = paste(str_remove(string = i, pattern = "[.]1"), "(ADT)")) +
      theme_bw() +
      theme(axis.text = element_text(size = 12),
            axis.title.x = element_text(size = 14, face = "plain"),
            axis.title.y = element_blank(),
            title = element_text(size = 16, face = "bold"),
            panel.grid = element_blank(),
            legend.text = element_text(size = 12),
            legend.position = "none") -> plot
  print(plot)
}
dev.off()

#- Proposed annotation
# 0 - EM CD4+ T (1)
# 1 - Naive-like CD4+ T (1)
# 2 - EM CD4+ T (1)
# 3 - Naive-like CD4+ T (1)
# 4 - CM CD4+ T
# 5 - EM CD4+ T (1)
# 6 - Early Differentiated EM CD4+ T
# 7 - EM CD4+ T (1)
# 8 - EM CD4+ T (1)
# 9 - ADT-high
# 10 - EM CD4+ T (2)
# 11 - Naive-like CD4+ T (1)
# 12 - EM CD4+ T (1)
# 13 - EM CD4+ T (1)
# 14 - Naive-like CD4+ T (2)
# 15 - EM CD4+ T (3)
# 16 - EM CD4+ T (4)
# 17 - CD103+ EM CD4+ T
# 18 - NKT & gdT
# 19 - Regulatory T

#- Add results to meta.data
integrated.seurat@meta.data$integrated.clusters <- integrated.seurat@meta.data$wsnn_res.1
integrated.seurat@meta.data$integrated.annotation.clusters <- plyr::mapvalues(x = integrated.seurat@meta.data$integrated.clusters, 
                                                                              from = 0:19, 
                                                                              to = c("EM CD4+ T (1)", 
                                                                                     "Naive-like CD4+ T (1)", 
                                                                                     "EM CD4+ T (1)", 
                                                                                     "Naive-like CD4+ T (1)", 
                                                                                     "CM CD4+ T",
                                                                                     "EM CD4+ T (1)", 
                                                                                     "Early Differentiated EM CD4+ T", 
                                                                                     "EM CD4+ T (1)", 
                                                                                     "EM CD4+ T (1)", 
                                                                                     "ADT-high", 
                                                                                     "EM CD4+ T (2)", 
                                                                                     "Naive-like CD4+ T (1)", 
                                                                                     "EM CD4+ T (1)", 
                                                                                     "EM CD4+ T (1)", 
                                                                                     "Naive-like CD4+ T (2)", 
                                                                                     "EM CD4+ T (3)", 
                                                                                     "EM CD4+ T (4)", 
                                                                                     "CD103+ EM CD4+ T", 
                                                                                     "NKT & gdT", 
                                                                                     "Regulatory T")) %>% as.character
integrated.seurat@meta.data$integrated.clusters <- plyr::mapvalues(x = integrated.seurat@meta.data$integrated.annotation.clusters,
                                                                   from = c("Naive-like CD4+ T (1)", 
                                                                            "Naive-like CD4+ T (2)", 
                                                                            "CM CD4+ T", 
                                                                            "Early Differentiated EM CD4+ T", 
                                                                            "EM CD4+ T (1)", 
                                                                            "EM CD4+ T (2)",
                                                                            "EM CD4+ T (3)", 
                                                                            "EM CD4+ T (4)", 
                                                                            "CD103+ EM CD4+ T",
                                                                            "Regulatory T",
                                                                            "NKT & gdT",
                                                                            "ADT-high"),
                                                                   to = paste("Seurat", 1:12))
integrated.seurat@meta.data$integrated.clusters <- factor(x = integrated.seurat@meta.data$integrated.clusters, levels = paste("Seurat", 1:12))
```
```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- Cluster annotation - GEX & ADT - version 2 (merged clusters)
Idents(integrated.seurat) <- integrated.seurat@meta.data$integrated.clusters
Idents(integrated.seurat) %>% table()

#- MAST (GEX)
# Perform differential expression for each of the merged clusters
DefaultAssay(integrated.seurat) <- "RNA"
markers.clusters_GEX_2 <- FindAllMarkers(object = integrated.seurat, 
                                         max.cells.per.ident = 5000,
                                         only.pos = TRUE, 
                                         test.use = "MAST", 
                                         latent.vars = c("nCount_RNA", "batch", "ptid", "arm", "stim", "visit"),
                                         verbose = TRUE)
markers.clusters_GEX_2 <- markers.clusters_GEX_2 %>%
  filter(p_val_adj < 1e-05) %>%
  arrange(cluster, p_val_adj, avg_log2FC) %>%
  left_join(y = unique(annotations[, c("gene_name", "description")]), by = c("gene" = "gene_name")) %>%
  filter(!str_detect(string = gene, pattern = "RP")) %>%
  mutate(`Annotation - Valentin` = plyr::mapvalues(x = cluster, 
                                                   to = c("Naive-like CD4+ T (1)", 
                                                          "Naive-like CD4+ T (2)", 
                                                          "CM CD4+ T", 
                                                          "Early Differentiated EM CD4+ T", 
                                                          "EM CD4+ T (1)", 
                                                          "EM CD4+ T (2)",
                                                          "EM CD4+ T (3)", 
                                                          "EM CD4+ T (4)", 
                                                          "CD103+ EM CD4+ T",
                                                          "Regulatory T",
                                                          "NKT & gdT",
                                                          "ADT-high"), 
                                                   from = paste("Seurat", 1:12))) 
markers.clusters_GEX_2 %>%
  write_csv(file = here("output", "002_output", "GEX_markers-clusters_v2.csv"))

#- Top 20 (GEX)
markers.clusters_GEX_2 %>%
  group_by(cluster) %>% 
  filter(!str_detect(string = gene, pattern = "RP")) %>%
  top_n(n = 20, wt = avg_log2FC) %>%
  View("Top genes")

#- LR (ADT)
# Perform differential expression for each of the merged clusters
DefaultAssay(integrated.seurat) <- "ADT"
markers.clusters_ADT_2 <- FindAllMarkers(object = integrated.seurat, 
                                         max.cells.per.ident = 5000,
                                         only.pos = TRUE, 
                                         test.use = "LR", 
                                         latent.vars = c("batch", "ptid", "arm", "stim", "visit"),
                                         verbose = TRUE)
markers.clusters_ADT_2 <- markers.clusters_ADT_2 %>%
  filter(p_val_adj < 1e-05) %>%
  arrange(cluster, p_val_adj, avg_log2FC) %>%
  mutate(`Annotation - Valentin` = plyr::mapvalues(x = cluster, 
                                                   to = c("Naive-like CD4+ T (1)", 
                                                          "Naive-like CD4+ T (2)", 
                                                          "CM CD4+ T", 
                                                          "Early Differentiated EM CD4+ T", 
                                                          "EM CD4+ T (1)", 
                                                          "EM CD4+ T (2)",
                                                          "EM CD4+ T (3)", 
                                                          "EM CD4+ T (4)", 
                                                          "CD103+ EM CD4+ T",
                                                          "Regulatory T",
                                                          "NKT & gdT",
                                                          "ADT-high"), 
                                                   from = paste("Seurat", 1:12))) 
markers.clusters_ADT_2 %>%
  write_csv(file = here("output", "002_output", "ADT_markers-clusters_v2.csv"))

#- Top 20 (ADT)
markers.clusters_ADT_2 %>%
  group_by(cluster) %>% 
  top_n(n = 20, wt = avg_log2FC) %>%
  View("Top ADTs")
```

According to these markers and after merging clusters that do not exhibit clear evidence of separation, we have  

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=11}
###--- UMAP
#- Data
dt.ggplot <- data.table(Embeddings(object = integrated.seurat, "umap_WNN"), integrated.seurat@meta.data)

#- Annotation
centers <- dt.ggplot %>% group_by(integrated.clusters, integrated.annotation.clusters) %>% summarize(UMAP_1 = median(wnnUMAP_1), UMAP_2 = median(wnnUMAP_2))

#- Plot
my.colors <- c(RColorBrewer::brewer.pal(n = 10, name = "Paired"), "gold2", "grey90")
names(my.colors) <- paste("Seurat", 1:12)
dt.ggplot %>%  
  filter(integrated.clusters %in% paste("Seurat", 1:12)) %>%
  ggplot(aes(x = wnnUMAP_1, y = wnnUMAP_2)) +
    geom_point(aes(color = integrated.clusters), size = .01) +
    labs(x = "UMAP-1", y = "UMAP-2", color = NULL) +
    scale_color_manual(values = my.colors) +
    geom_text(aes(x = UMAP_1, y = UMAP_2, label = integrated.annotation.clusters), size = 4, color = "black", data = centers, fontface = "bold") +
    guides(colour = guide_legend(override.aes = list(size = 4, alpha = 1))) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          axis.title = element_text(size = 14),
          panel.grid = element_blank(),
          legend.text = element_text(size = 12),
          legend.position = "bottom")
```

**Heatmap - GEX**

Not shown.  

```{r echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
###--- Heatmap
#- Top 10
DEGs <- markers.clusters_GEX_2 %>%
  filter(cluster %in% paste("Seurat", 1:11)) %>%
  group_by(cluster) %>% 
  filter(!str_detect(string = gene, pattern = "RP")) %>%
  top_n(n = 10, wt = avg_log2FC) %>%
  pull(gene)

#- Heatmap (downsample - Seurat)
my.colors <- c(RColorBrewer::brewer.pal(n = 10, name = "Paired"), "gold2")
names(my.colors) <- paste("Seurat", 1:11)

DefaultAssay(integrated.seurat) <- "RNA"
seurat.tmp <- subset(integrated.seurat, integrated.clusters %in% paste("Seurat", 1:11))
Idents(seurat.tmp) <- factor(x = seurat.tmp@meta.data$integrated.clusters, levels = paste("Seurat", 1:11)) 
seurat.tmp <- subset(seurat.tmp, downsample = 100)
seurat.tmp <- ScaleData(object = seurat.tmp, features = DEGs, verbose = TRUE)

png(filename = here("output", "002_output", "GEX_heatmap-top10_v2.png"), width = 1000, height = 1300, bg = "white")
DoHeatmap(seurat.tmp, features = DEGs, group.colors = my.colors)
dev.off()
```

**Ridge plots - ADT**

Not shown.  

```{r echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
###--- RidgePlot - ADTs
my.colors <- c(RColorBrewer::brewer.pal(n = 10, name = "Paired"), "gold2")
names(my.colors) <- paste("Seurat", 1:11)

pdf(file = here("output", "002_output", "ADT_ridgePlots_v2.pdf"), width = 5, height = 8, bg = "white", pointsize = 12)
for(i in setdiff(rownames(integrated.seurat@assays$ADT@data), "CD4.2")){
  data.table(integrated.seurat@meta.data, ADT = integrated.seurat@assays$ADT@data[i, ]) %>%
    filter(integrated.clusters %in% paste("Seurat", 1:11)) %>%
    mutate(integrated.clusters = factor(x = integrated.clusters, levels = paste("Seurat", 1:11))) %>%
    ggplot(aes(y = integrated.clusters, x = ADT, fill = integrated.clusters)) +
      ggridges::geom_density_ridges(bandwidth = .2) +
      geom_vline(xintercept = 0, color = "grey20", alpha = .7, lty = 2) +
      labs(x = "Normalized ADT expression", title = paste(str_remove(string = i, pattern = "[.]1"), "(ADT)")) +
      scale_fill_manual(values = my.colors) +
      theme_bw() +
      theme(axis.text = element_text(size = 12),
            axis.title.x = element_text(size = 14, face = "plain"),
            axis.title.y = element_blank(),
            title = element_text(size = 16, face = "bold"),
            panel.grid = element_blank(),
            legend.text = element_text(size = 12),
            legend.position = "none") -> plot
  print(plot)
}
dev.off()
```

## Comparison: Seurat 1 vs Seurat 2

Seurat 1 and Seurat 2 are compared to each other as they are both annotated as being naive-like CD4+ T - see below. 

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- Seurat 1 vs Seurat 2 - GEX & ADT
Idents(integrated.seurat) <- integrated.seurat@meta.data$integrated.clusters
Idents(integrated.seurat) %>% table()

#- MAST (GEX)
DefaultAssay(integrated.seurat) <- "RNA"
markers.clusters_GEX_1vs2 <- FindMarkers(object = integrated.seurat, 
                                         ident.1 = "Seurat 1", 
                                         ident.2 = "Seurat 2", 
                                         test.use = "MAST", 
                                         latent.vars = c("nCount_RNA", "batch", "ptid", "arm", "stim", "visit"),
                                         verbose = TRUE)
markers.clusters_GEX_1vs2 <- markers.clusters_GEX_1vs2 %>%
  rownames_to_column("gene") %>%
  filter(p_val_adj < .01) %>%
  filter(abs(avg_log2FC) >= log2(1.5)) %>%
  arrange(p_val_adj, avg_log2FC) %>%
  left_join(y = unique(annotations[, c("gene_name", "description")]), by = c("gene" = "gene_name")) %>%
  mutate(contrast = "Seurat 1 vs Seurat 2") %>%
  select(gene, contrast, avg_log2FC, pct.1, pct.2, p_val_adj, p_val, description)
markers.clusters_GEX_1vs2 %>%
  arrange(desc(avg_log2FC)) %>%
  write_csv(file = here("output", "002_output", "GEX_markers-clusters_Seurat1-vs-Seurat2_version2.csv"))
View(markers.clusters_GEX_1vs2)

#- LR (ADT)
DefaultAssay(integrated.seurat) <- "ADT"
markers.clusters_ADT_1vs2 <- FindMarkers(object = integrated.seurat, 
                                         ident.1 = "Seurat 1", 
                                         ident.2 = "Seurat 2", 
                                         test.use = "LR", 
                                         latent.vars = c("batch", "ptid", "arm", "stim", "visit"),
                                         verbose = TRUE)
markers.clusters_ADT_1vs2 <- markers.clusters_ADT_1vs2 %>%
  rownames_to_column("ADT") %>%
  filter(p_val_adj < .1) %>%
  arrange(p_val_adj, avg_log2FC) %>%
  mutate(contrast = "Seurat 1 vs Seurat 2") %>%
  select(ADT, contrast, avg_log2FC, pct.1, pct.2, p_val, p_val_adj)
markers.clusters_ADT_1vs2 %>%
  write_csv(file = here("output", "002_output", "ADT_markers-clusters_Seurat1-vs-Seurat2.csv"))
View(markers.clusters_ADT_1vs2)
```

## Comparison: Seurat 5 vs Seurat 6 vs Seurat 7 vs Seurat 8

Seurat 5, 6, 7 and 8 are compared to each other as they are annotated as being EM CD4+ T - see below. 

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- Seurat 5, 6, 7 and 8 - GEX & ADT
Idents(integrated.seurat) <- integrated.seurat@meta.data$integrated.clusters
Idents(integrated.seurat) %>% table()

#- MAST (GEX)
markers.clusters_GEX_EM <- foreach(i = 1:4) %do% 
{
  ident.1 <- paste("Seurat", 5:8)[i]
  ident.2 <- list(c("Seurat 6", "Seurat 7", "Seurat 8"),
                  c("Seurat 5", "Seurat 7", "Seurat 8"),
                  c("Seurat 5", "Seurat 6", "Seurat 8"),
                  c("Seurat 5", "Seurat 6", "Seurat 7"))[[i]]
  contrast <- c("Seurat 5 vs Seurat 6, 7 & 8",
                "Seurat 6 vs Seurat 5, 7 & 8",
                "Seurat 7 vs Seurat 5, 6 & 8",
                "Seurat 8 vs Seurat 5, 6 & 7")[i]
  DefaultAssay(integrated.seurat) <- "RNA"
  markers.clusters_GEX_EM.tmp <- FindMarkers(object = integrated.seurat, 
                                             ident.1 = ident.1, 
                                             ident.2 = ident.2, 
                                             test.use = "MAST", 
                                             latent.vars = c("nCount_RNA", "batch", "ptid", "arm", "stim", "visit"),
                                             verbose = TRUE)
  markers.clusters_GEX_EM.tmp %>%
    rownames_to_column("gene") %>%
    filter(p_val_adj < .01) %>%
    filter(abs(avg_log2FC) >= log2(1.5)) %>%
    arrange(p_val_adj, avg_log2FC) %>%
    left_join(y = unique(annotations[, c("gene_name", "description")]), by = c("gene" = "gene_name")) %>%
    mutate(contrast = contrast) %>%
    select(gene, contrast, avg_log2FC, pct.1, pct.2, p_val_adj, description) %>%
    return()
}
markers.clusters_GEX_EM <- bind_rows(markers.clusters_GEX_EM)
markers.clusters_GEX_EM %>%
  write_csv(file = here("output", "002_output", "GEX_markers-clusters_Seurat5-Seurat6-Seurat7-Seurat8.csv"))
View(markers.clusters_GEX_EM)

#- LR (ADT)
markers.clusters_ADT_EM <- foreach(i = 1:4) %do% 
{
  ident.1 <- paste("Seurat", 5:8)[i]
  ident.2 <- list(c("Seurat 6", "Seurat 7", "Seurat 8"),
                  c("Seurat 5", "Seurat 7", "Seurat 8"),
                  c("Seurat 5", "Seurat 6", "Seurat 8"),
                  c("Seurat 5", "Seurat 6", "Seurat 7"))[[i]]
  contrast <- c("Seurat 5 vs Seurat 6, 7 & 8",
                "Seurat 6 vs Seurat 5, 7 & 8",
                "Seurat 7 vs Seurat 5, 6 & 8",
                "Seurat 8 vs Seurat 5, 6 & 7")[i]
  DefaultAssay(integrated.seurat) <- "ADT"
  markers.clusters_ADT_EM.tmp <- FindMarkers(object = integrated.seurat, 
                                             ident.1 = ident.1, 
                                             ident.2 = ident.2, 
                                             test.use = "LR", 
                                             latent.vars = c("batch", "ptid", "arm", "stim", "visit"),
                                             verbose = TRUE)
  markers.clusters_ADT_EM.tmp %>%
    rownames_to_column("ADT") %>%
    filter(p_val_adj < .1) %>%
    arrange(p_val_adj, avg_log2FC) %>%
    mutate(contrast = contrast) %>%
    select(ADT, contrast, avg_log2FC, pct.1, pct.2, p_val_adj) %>%
    return()
}
markers.clusters_ADT_EM <- bind_rows(markers.clusters_ADT_EM)
markers.clusters_ADT_EM %>%
  write_csv(file = here("output", "002_output", "ADT_markers-clusters_Seurat5-Seurat6-Seurat7-Seurat8.csv"))
View(markers.clusters_ADT_EM)
```
```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- Seurat 5, 6, 7 and 8 - GEX & ADT - version 2
Idents(integrated.seurat) <- integrated.seurat@meta.data$integrated.clusters
Idents(integrated.seurat) %>% table()

#- MAST (GEX)
markers.clusters_GEX_EM_2 <- foreach(i = 1:6) %do% 
{
  ident.1 <- c("Seurat 5", "Seurat 5", "Seurat 5", "Seurat 6", "Seurat 6", "Seurat 7")[i]
  ident.2 <- c("Seurat 6", "Seurat 7", "Seurat 8", "Seurat 7", "Seurat 8", "Seurat 8")[i]
  contrast <- c("Seurat 5 vs Seurat 6",
                "Seurat 5 vs Seurat 7",
                "Seurat 5 vs Seurat 8",
                "Seurat 6 vs Seurat 7",
                "Seurat 6 vs Seurat 8",
                "Seurat 7 vs Seurat 8")[i]
  DefaultAssay(integrated.seurat) <- "RNA"
  markers.clusters_GEX_EM.tmp <- FindMarkers(object = integrated.seurat, 
                                             ident.1 = ident.1, 
                                             ident.2 = ident.2, 
                                             test.use = "MAST", 
                                             latent.vars = c("nCount_RNA", "batch", "ptid", "arm", "stim", "visit"),
                                             verbose = TRUE)
  markers.clusters_GEX_EM.tmp %>%
    rownames_to_column("gene") %>%
    filter(p_val_adj < .01) %>%
    filter(abs(avg_log2FC) >= log2(1.5)) %>%
    arrange(p_val_adj, avg_log2FC) %>%
    left_join(y = unique(annotations[, c("gene_name", "description")]), by = c("gene" = "gene_name")) %>%
    mutate(contrast = contrast) %>%
    select(gene, contrast, avg_log2FC, pct.1, pct.2, p_val_adj, description) %>%
    return()
}
markers.clusters_GEX_EM_2 <- bind_rows(markers.clusters_GEX_EM_2)
markers.clusters_GEX_EM_2 %>%
  write_csv(file = here("output", "002_output", "GEX_markers-clusters_Seurat5-Seurat6-Seurat7-Seurat8_v2.csv"))
View(markers.clusters_GEX_EM_2)

#- LR (ADT)
markers.clusters_ADT_EM_2 <- foreach(i = 1:6) %do% 
{
  ident.1 <- c("Seurat 5", "Seurat 5", "Seurat 5", "Seurat 6", "Seurat 6", "Seurat 7")[i]
  ident.2 <- c("Seurat 6", "Seurat 7", "Seurat 8", "Seurat 7", "Seurat 8", "Seurat 8")[i]
  contrast <- c("Seurat 5 vs Seurat 6",
                "Seurat 5 vs Seurat 7",
                "Seurat 5 vs Seurat 8",
                "Seurat 6 vs Seurat 7",
                "Seurat 6 vs Seurat 8",
                "Seurat 7 vs Seurat 8")[i]
  DefaultAssay(integrated.seurat) <- "ADT"
  markers.clusters_ADT_EM.tmp <- FindMarkers(object = integrated.seurat, 
                                             ident.1 = ident.1, 
                                             ident.2 = ident.2, 
                                             test.use = "LR", 
                                             latent.vars = c("batch", "ptid", "arm", "stim", "visit"),
                                             verbose = TRUE)
  markers.clusters_ADT_EM.tmp %>%
    rownames_to_column("ADT") %>%
    filter(p_val_adj < .1) %>%
    arrange(p_val_adj, avg_log2FC) %>%
    mutate(contrast = contrast) %>%
    select(ADT, contrast, avg_log2FC, pct.1, pct.2, p_val_adj) %>%
    return()
}
markers.clusters_ADT_EM_2 <- bind_rows(markers.clusters_ADT_EM_2)
markers.clusters_ADT_EM_2 %>%
  write_csv(file = here("output", "002_output", "ADT_markers-clusters_Seurat5-Seurat6-Seurat7-Seurat8_v2.csv"))
View(markers.clusters_ADT_EM_2)
```

## Comparison: Seurat 9 vs Seurat 5 or Seurat 6 or Seurat 7 or Seurat 8  

Seurat 9 and Seurat 5, 6, 7 or 8 are compared to each other - see below.  

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- Seurat 9 vs 5, 6, 7 and 8 - GEX & ADT
Idents(integrated.seurat) <- integrated.seurat@meta.data$integrated.clusters
Idents(integrated.seurat) %>% table()

#- MAST (GEX)
markers.clusters_GEX_Seurat9 <- foreach(i = 1:4) %do% 
{
  ident.1 <- "Seurat 9"
  ident.2 <- c("Seurat 5", "Seurat 6", "Seurat 7", "Seurat 8")[i]
  contrast <- c("Seurat 9 vs Seurat 5",
                "Seurat 9 vs Seurat 6",
                "Seurat 9 vs Seurat 7",
                "Seurat 9 vs Seurat 8")[i]
  DefaultAssay(integrated.seurat) <- "RNA"
  markers.clusters_GEX.tmp <- FindMarkers(object = integrated.seurat, 
                                          ident.1 = ident.1, 
                                          ident.2 = ident.2, 
                                          test.use = "MAST", 
                                          latent.vars = c("nCount_RNA", "batch", "ptid", "arm", "stim", "visit"),
                                          verbose = TRUE)
  markers.clusters_GEX.tmp %>%
    rownames_to_column("gene") %>%
    filter(p_val_adj < .01) %>%
    filter(abs(avg_log2FC) >= log2(1.5)) %>%
    arrange(p_val_adj, avg_log2FC) %>%
    left_join(y = unique(annotations[, c("gene_name", "description")]), by = c("gene" = "gene_name")) %>%
    mutate(contrast = contrast) %>%
    select(gene, contrast, avg_log2FC, pct.1, pct.2, p_val_adj, description) %>%
    return()
}
markers.clusters_GEX_Seurat9 <- bind_rows(markers.clusters_GEX_Seurat9)
markers.clusters_GEX_Seurat9 %>%
  write_csv(file = here("output", "002_output", "GEX_markers-clusters_Seurat9.csv"))
View(markers.clusters_GEX_Seurat9)

#- LR (ADT)
markers.clusters_ADT_Seurat9 <- foreach(i = 1:4) %do% 
{
  ident.1 <- "Seurat 9"
  ident.2 <- c("Seurat 5", "Seurat 6", "Seurat 7", "Seurat 8")[i]
  contrast <- c("Seurat 9 vs Seurat 5",
                "Seurat 9 vs Seurat 6",
                "Seurat 9 vs Seurat 7",
                "Seurat 9 vs Seurat 8")[i]
  DefaultAssay(integrated.seurat) <- "ADT"
  markers.clusters_ADT_EM.tmp <- FindMarkers(object = integrated.seurat, 
                                             ident.1 = ident.1, 
                                             ident.2 = ident.2, 
                                             test.use = "LR", 
                                             latent.vars = c("batch", "ptid", "arm", "stim", "visit"),
                                             verbose = TRUE)
  markers.clusters_ADT_EM.tmp %>%
    rownames_to_column("ADT") %>%
    filter(p_val_adj < .1) %>%
    arrange(p_val_adj, avg_log2FC) %>%
    mutate(contrast = contrast) %>%
    select(ADT, contrast, avg_log2FC, pct.1, pct.2, p_val_adj) %>%
    return()
}
markers.clusters_ADT_Seurat9 <- bind_rows(markers.clusters_ADT_Seurat9)
markers.clusters_ADT_Seurat9 %>%
  write_csv(file = here("output", "002_output", "ADT_markers-clusters_Seurat9.csv"))
View(markers.clusters_ADT_Seurat9)
```

## Comparison: Seurat 4 vs Seurat 5 vs Seurat 6 vs Seurat 7 vs Seurat 9

Seurat 4, 5, 6, 7, 8 and 9 are compared to each other as they are annotated as being EM CD4+ T - see below. 

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- Seurat 4, 5, 6, 7, 8 & 9 - GEX & ADT
integrated.seurat.tmp <- subset(integrated.seurat, integrated.clusters %in% paste("Seurat", 4:9))
Idents(integrated.seurat.tmp) <- integrated.seurat.tmp@meta.data$integrated.clusters
Idents(integrated.seurat.tmp) %>% table()

###########
#- MAST (GEX)
DefaultAssay(integrated.seurat.tmp) <- "RNA"
markers.clusters_GEX_EM_3 <- FindAllMarkers(object = integrated.seurat.tmp, 
                                            only.pos = TRUE, 
                                            test.use = "MAST", 
                                            latent.vars = c("nCount_RNA", "batch", "ptid", "arm", "stim", "visit"),
                                            verbose = TRUE)
markers.clusters_GEX_EM_3 <- markers.clusters_GEX_EM_3 %>%
  filter(p_val_adj < 1e-05) %>%
  arrange(cluster, p_val_adj, avg_log2FC) %>%
  left_join(y = unique(annotations[, c("gene_name", "description")]), by = c("gene" = "gene_name")) %>%
  filter(!str_detect(string = gene, pattern = "RP")) %>%
  mutate(`Annotation - Valentin` = plyr::mapvalues(x = cluster, 
                                                   to = c("Early Differentiated EM CD4+ T", 
                                                          "EM CD4+ T (1)", 
                                                          "EM CD4+ T (2)",
                                                          "EM CD4+ T (3)", 
                                                          "EM CD4+ T (4)", 
                                                          "CD103+ EM CD4+ T"), 
                                                   from = paste("Seurat", 4:9))) 
markers.clusters_GEX_EM_3 %>%
  write_csv(file = here("output", "002_output", "GEX_markers-clusters_Seurat4-Seurat5-Seurat6-Seurat7-Seurat8-Seurat9.csv"))

#- Top 20 (GEX)
markers.clusters_GEX_EM_3 %>%
  group_by(cluster) %>% 
  filter(!str_detect(string = gene, pattern = "RP")) %>%
  top_n(n = 20, wt = avg_log2FC) %>%
  View("Top genes")

###########
#- LR (ADT)
DefaultAssay(integrated.seurat.tmp) <- "ADT"
markers.clusters_ADT_EM_3 <- FindAllMarkers(object = integrated.seurat.tmp, 
                                            only.pos = TRUE, 
                                            test.use = "LR", 
                                            latent.vars = c("batch", "ptid", "arm", "stim", "visit"),
                                            verbose = TRUE)
markers.clusters_ADT_EM_3 <- markers.clusters_ADT_EM_3 %>%
  filter(p_val_adj < 1e-05) %>%
  arrange(cluster, p_val_adj, avg_log2FC) %>%
  mutate(`Annotation - Valentin` = plyr::mapvalues(x = cluster, 
                                                   to = c("Early Differentiated EM CD4+ T", 
                                                          "EM CD4+ T (1)", 
                                                          "EM CD4+ T (2)",
                                                          "EM CD4+ T (3)", 
                                                          "EM CD4+ T (4)", 
                                                          "CD103+ EM CD4+ T"), 
                                                   from = paste("Seurat", 4:9))) 
markers.clusters_ADT_EM_3 %>%
  write_csv(file = here("output", "002_output", "ADT_markers-clusters_Seurat4-Seurat5-Seurat6-Seurat7-Seurat8-Seurat9.csv"))

#- Top 20 (ADT)
markers.clusters_ADT_EM_3 %>%
  group_by(cluster) %>% 
  top_n(n = 20, wt = avg_log2FC) %>%
  View("Top ADTs")
```

